Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-06 treestats_paper.tex
Comment re: missing refs
Mention speed+RAM in abstract
Comment on 1st sentence

@@

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
thus providing a glimpse of the genealogical tree that relates the genomes to each other at that site.
Summaries of genetic variation thus also aim to summarize the underlying genealogies.
Here we exploit this duality to create a method that efficiently computes single-site statistics
from population genetic data using the succinct tree sequence encoding
of genome sequence and genealogies.
We use the framework to provide fast implementations of many currently-defined statistics
(whose run time scales with the logarithm of the number of samples),
and to suggest several statistics.
The general approach aggregates ``sample weights'' up each marginal tree,
which are then summarised using a ``summary function'';
different statistics result from different choices of weight or function.
The results can be summarised in three different ways:
by \emph{site}, which corresponds to statistics calculated from genome sequence,
by \emph{branch}, which gives the expected value of the site statistic
with a large number of infinite-sites mutations,
or by \emph{node}, which summarises the contribution of each node in the tree to the statistics.
This formulation also makes it easy to see what statistic of the underlying genealogical trees
each sequence-based statistic is estimating.

\plr{add bits for speedup; data}

@@

It was once a major undertaking to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep{stankowski2019widespread,fst_landscapes},
or between individuals rather than populations \citep{individual_stats}.
Computation is beginning to be a major problem:
whole-genome sequence on the scale of the UK Biobank,
for instance, might have 500K individuals genotyped at 10M loci \citep{ukbb},
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Indeed, many sorts of modern inference machinery, such as deep learning,
depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.

>>>>

\krt{several refs seem missing}

@@

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
thus providing a glimpse of the genealogical tree that relates the genomes to each other at that site.
Summaries of genetic variation thus also aim to summarize the underlying genealogies.
Here we exploit this duality to create a method that efficiently computes single-site statistics
from population genetic data using the succinct tree sequence encoding
of genome sequence and genealogies.
We use the framework to provide fast implementations of many currently-defined statistics
(whose run time scales with the logarithm of the number of samples),
and to suggest several statistics.
The general approach aggregates ``sample weights'' up each marginal tree,
which are then summarised using a ``summary function'';
different statistics result from different choices of weight or function.
The results can be summarised in three different ways:
by \emph{site}, which corresponds to statistics calculated from genome sequence,
by \emph{branch}, which gives the expected value of the site statistic
with a large number of infinite-sites mutations,
or by \emph{node}, which summarises the contribution of each node in the tree to the statistics.
This formulation also makes it easy to see what statistic of the underlying genealogical trees
each sequence-based statistic is estimating.  By using simulations to evaluate performance, we 
show that calculating statistics from tree sequences is several orders of magnitude more efficient than
methods traversing the entire data matrix in terms of both run time and memory requirements.

\plr{add bits for data}

@@

\krt{Why? I think this is an odd way of putting it.  The challenge back then was
    genotyping.  Once that was done, summarizing the relationships was done on a 
    calculator b/c the data sets were so small.}
It was once a major undertaking to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep{stankowski2019widespread,fst_landscapes},
or between individuals rather than populations \citep{individual_stats}.
Computation is beginning to be a major problem:
whole-genome sequence on the scale of the UK Biobank,
for instance, might have 500K individuals genotyped at 10M loci \citep{ukbb},
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Indeed, many sorts of modern inference machinery, such as deep learning,
depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.
